---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r get&clean}
unzip("activity.zip")
data <- read.csv("activity.csv")
library(dplyr, warn.conflicts = FALSE)
data_day <- data %>% group_by(date) %>% summarize(total_steps = sum(steps))
data_interval <- data %>% group_by(interval) %>% 
        summarize(mean_steps = mean(steps, na.rm=TRUE))
```


## What is mean total number of steps taken per day?
```{r steps_per_day}
with(data_day, hist(total_steps, col="red", xlab="total steps", ylim=c(0,30), 
     main ="total steps per day"))

total_steps_mean <- mean(data_day$total_steps, na.rm=TRUE)
total_steps_median <- median(data_day$total_steps, na.rm=TRUE)
```

**total steps mean:** `r format(total_steps_mean, scientific=FALSE)`  
**total steps median:** `r total_steps_median`  
  

## What is the average daily activity pattern?
```{r steps_per_interval}
with(data_interval, plot(interval, mean_steps, type="l", col="blue", 
                         xlab="interval", ylab="average steps"))
title(main ="average steps per interval")

max_interval <- data_interval %>% filter(mean_steps==max(mean_steps))
```

**maximum steps interval =** `r max_interval[[1,1]]`  
  

## Imputing missing values
### Total number of missing values
```{r missing_values}
sum(is.na(data$steps))
```
### New data with missing values filled in with the mean for that interval
```{r new_data}
ndata <- merge(data, data_interval, by.x="interval", by.y="interval", all=TRUE)
x <- is.na(ndata$steps)
ndata$steps[x] <- ndata$mean_steps[x]
```
### Histogram of total steps by day, mean and median
```{r new_steps_per_day}
ndata_day <- ndata %>% group_by(date) %>% summarize(total_steps = sum(steps))

with(ndata_day, hist(total_steps, col="red", xlab="total steps", ylim=c(0,35), 
     main ="total steps per day"))

ntotal_steps_mean <- mean(ndata_day$total_steps, na.rm=TRUE)
ntotal_steps_median <- median(ndata_day$total_steps, na.rm=TRUE)
```

**new total steps mean:** `r format(ntotal_steps_mean, scientific=FALSE)`  
**new total steps median:** `r format(ntotal_steps_median, scientific=FALSE)`  
Both values do not barely differ from first estimated values.  

## Are there differences in activity patterns between weekdays and weekends?
```{r steps_per_interval&daytype}
Sys.setlocale("LC_TIME", "English")
ndata$daytype <- factor(ifelse(weekdays(as.Date(ndata$date)) %in% 
                                       c("Saturday", "Sunday"), "weekend", "weekday"))
ndata_interval_wd <- ndata %>% group_by(interval, daytype) %>%
        summarize(nmean_steps=mean(steps))

library(ggplot2, warn.conflicts = FALSE)
g <- ggplot(ndata_interval_wd, aes(interval, nmean_steps))
g + geom_line() + facet_grid(daytype~.) + 
        labs(title="Average steps per interval and day type", y="average steps")
```
  
There are visible differences between weekdays and weekends. On weekends the distribution
is more uniform during the day, and the maximum at 08:35 corresponds to weekdays.  


